---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


The following function **assumes dplyr and ggplot2 are installed** and loads them if not already loaded.  
This function was adapted from maloneypatr's solution on [this web page](http://stackoverflow.com/questions/15155814/check-if-r-package-is-installed-then-load-library) at stackoverflow.

```{r loadPackages, message=FALSE}
load_pkgs <- function(pkg1, ...)  {
     
      # convert arguments to a vector
      packages <- c(pkg1, ...)
     
      # start loop to determine if each package is loaded
      for(package in packages)  {
            
            # if package is not loaded, load
            if(!package %in% search())  {
                  do.call("library", list(package))
                  }
            }
      }
load_pkgs(c("dplyr", "ggplot2"))
```            


Per course instructions the "activity.zip" dataset is stored locally.  
Extract the dataset if needed, then format and read. 
```{r extractAndRead}
if(!file.exists("activity.csv"))  {
      unzip("activity.zip", exdir = ".")
      }
dat <- tbl_df(read.csv("activity.csv"))
```


Dataset structure
```{r} 
glimpse(dat)
head(dat, 3)
tail(dat, 3)

summary(dat)
```


Coerce "date" variable
```{r}
dat$date <- as.Date(dat$date)
```


## What is mean total number of steps taken per day?  

Ignore missing values for this part of the assignment (per instructions)

#### 1. Calculate total number of steps taken per day
```{r}
totalSteps <- dat %>%
      group_by(date) %>%
      summarise(sums = sum(steps))
```


#### 2. Histogram of total number of steps per day
```{r, totalSteps1, fig.align='left', fig.height=5, message=FALSE}
ggplot(totalSteps, aes(x=sums)) +
      geom_histogram(aes(fill = ..count..), binwidth = 1000) +
      geom_text(data = NULL, x = 17500, y = 8.2,
                label = "bin width = 1000", colour = "darkgray") +
      xlab("total steps") +
      ggtitle("Histogram of total steps per day")
```


#### 3. Mean and median of total number of steps per day
```{r}
mean(na.omit(totalSteps$sums))
median(na.omit(totalSteps$sums))
```


## What is the average daily activity pattern?

#### 1. Time series plot of the 5-minute interval and average number of steps taken, averaged across all days
```{r int_averages}
int_averages <- dat %>%
      group_by(interval) %>%
      summarise(mean_steps = mean(na.omit(steps)))

head(int_averages, 3)
```

Time series plot of mean of intervals
```{r, IntervalAvgs, fig.align='left', fig.height=5, message=FALSE}
ggplot(int_averages, aes(x=interval, y=mean_steps)) +
      geom_line() +
      xlab("interval id") +
      ylab("average steps") +
      ggtitle("Average steps per daily 5 minute interval")
```


#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxInt <- int_averages %>%
      filter(mean_steps == max(mean_steps))

maxInt
```

Interval id #`r maxInt$interval` contains the maximum average number of steps, approximately `r round(maxInt$mean_steps, 0)`. 

## Imputing missing values

#### 1. Total number of missing values in the dataset 
```{r}
length(which(complete.cases(dat) == FALSE))
```

`r length(which(complete.cases(dat) == FALSE))` 
of the `r nrow(dat)` rows are missing values.  

#### 2. Strategy for filling in missing values:  
Replace missing values with the mean calculated for each interval.

#### 3. New dataset with imputed values
```{r}
dat2 <- dat
for(i in 1:nrow(dat2))  {
      dat2[i, 1] <- ifelse(
            is.na(dat2$steps[i]),
            int_averages[which(int_averages$interval == dat2$interval[i]), "mean_steps"],
            dat2$steps[i])
            }
```


#### 4. New histogram with imputed values
```{r}
totalSteps2 <- dat2 %>%
      group_by(date) %>%
      summarise(sums = sum(steps))
```

```{r, totalSteps2, fig.align='left', fig.height=5, message=FALSE}
ggplot(totalSteps2, aes(x=sums)) +
      geom_histogram(aes(fill = ..count..), binwidth = 1000) +
      geom_text(data = NULL, x = 17500, y = 13.2, colour = "darkgray",
                label = "bin width = 1000") +
      xlab("total steps") +
      ggtitle("Histogram of total steps per day with imputed values")
```


##### 4.1 Mean and median of new dataset with imputed values
```{r}
mean(na.omit(totalSteps2$sums))
median(na.omit(totalSteps2$sums))
```


#### 4.2 Differences in mean and median: new vs. old
```{r}
x <- mean(na.omit(totalSteps2$sums)) - mean(na.omit(totalSteps$sums))
y <- median(na.omit(totalSteps2$sums)) - median(na.omit(totalSteps$sums))
x ; y
```

### Conclusion: 
When missing values are imputed the difference between the two datasets in mean (`r x`) and median (`r round(y, 0)`) appears insignificant. Overlaid histogram and density plots (below) show that imputation heightens and sharpens kurtosis.  


```{r}

########  Please note these combined overlaid plots    ########              
########  were not required for the assignement; they  ########       
########  were added for visualizing the comparison.  ######## 

# Visual comparison of the two datasets  
# Overlay histogram and density plots

# combine dataframes into one after adding a new variable
# to distinguish which total steps are from which

totalSteps3 <- totalSteps ; totalSteps4 <- totalSteps2

totalSteps3$dataset <- "old"
totalSteps4$dataset <- "new"

totalSteps5 <- rbind(totalSteps3, totalSteps4)
```

```{r, overlaidDensity, fig.align='left', message=FALSE, warning=FALSE}

# These plots are better stacked on each other, so prepare
# for processing via function shown further below

p1 <- ggplot(totalSteps5, aes(sums, fill = dataset)) +
      geom_histogram(alpha = 0.5, aes(y = ..density..),
                     position = "identity", binwidth = 1000) +
      scale_fill_discrete(
            breaks = c("new", "old"),
            labels = c("imputed", "omitted")) +
      theme(legend.position = c(1, 1), legend.justification = c(1,1)) +
      theme(legend.background = element_blank()) +
      theme(legend.key = element_blank()) +
      theme(axis.ticks = element_blank(), axis.text.x = element_blank()) +
      theme(axis.title.x=element_blank()) +
      ggtitle("Total steps per day: imputed vs. missing data omitted")

p2 <- ggplot(totalSteps5, aes(sums, fill = dataset)) +
      geom_density(alpha = 0.2) +
      theme(legend.position = "none") +
      xlab("total steps per day")

# Function for arranging ggplots in one window (from S. Turner at
# http://www.gettinggeneticsdone.com/2010/03/arrange-multiple-ggplot2-plots-in-same.html)

require(grid)
vp.layout <- function(x, y) viewport(layout.pos.row=x, layout.pos.col=y)
arrange_ggplot2 <- function(..., nrow=NULL, ncol=NULL, as.table=FALSE) {
  dots <- list(...)
  n <- length(dots)
  if(is.null(nrow) & is.null(ncol)) { nrow = floor(n/2) ; ncol = ceiling(n/nrow)}
  if(is.null(nrow)) { nrow = ceiling(n/ncol)}
  if(is.null(ncol)) { ncol = ceiling(n/nrow)}
  ## NOTE see n2mfrow in grDevices for possible alternative
  grid.newpage()
  pushViewport(viewport(layout=grid.layout(nrow,ncol) ) )
  ii.p <- 1
  for(ii.row in seq(1, nrow)){
    ii.table.row <- ii.row
    if(as.table) {ii.table.row <- nrow - ii.table.row + 1}
    for(ii.col in seq(1, ncol)){
      ii.table <- ii.p
      if(ii.p > n) break
      print(dots[[ii.table]], vp=vp.layout(ii.table.row, ii.col))
      ii.p <- ii.p + 1
    }
  }
}
arrange_ggplot2(p1, p2, nrow = 2)
```


## Are there differences in activity patterns between weekdays and weekends?

#### 1. Using the new dataset create a new factor variable with two levels - "weekday" and "weekend"  
```{r}
dat3 <- dat2
days <- lapply(dat2$date, weekdays)
dat3$date_type  <- as.factor(ifelse(grepl("Sat|Sun", days), "weekend", "weekday"))

# I prefer this order of columns 
dat3 <- dat3 %>%
select(steps, date, date_type, interval)
```


#### 2. Panel plot of average steps per 5-minute intervals: one for weekend days, the other for weekday days.
```{r intAveragesFactored}
int_averages2 <- dat3 %>%
      group_by(date_type, interval) %>%
      summarise(mean_steps = mean(steps))
```


Time series panel plot of mean of intervals
```{r, WeekendsWeekdays, fig.align='left', message=FALSE}
ggplot(int_averages2, aes(x=interval, y=mean_steps)) +
      geom_line() +
      facet_grid(date_type ~ .) +
      xlab("interval id") +
      ylab("average steps") +
      ggtitle("Average steps per daily 5 minute interval")
```


### Conclusion:  
There is an apparent difference in weekend day and weekday day interval averages. On weekdays activity sharply peaks in the early afternoon (around 2 pm). This is somewhat less apparent on weekends, which also show more activity in the later afternoon and evening than weekdays.